\subsection{Conditional Gaussian distributions/incremental simulation}
\label{sec: conditional gaussian distribution}

A Gaussian random function \(\rf\) has the property that for any finite
set of evaluation points \(x_1, \dots, x_n\) we have that
\[
    \rf(x_1), \dots, \rf(x_n)
\]
is a multivariate Gaussian random vector with mean
\[
    \mu(x_1), \dots, \mu(x_n)
\]
and covariance
\[
    \Sigma = \begin{pmatrix}
        \C(x_1, x_1) & \cdots & \C(x_1, x_n)
        \\
        \vdots & & \vdots
        \\
        \C(x_n, x_1) & \cdots & \C(x_n, x_n)
    \end{pmatrix}
\]
\(\mu\) is a ``mean function'' and \(\C\) a ``covariance function''.
To simulate \(\rf\) we essentially allow the user to evaluate this incrementally.
If the user has already requested the values \(\rf(x_1), \dots, \rf(x_n)\)
and then requests \(\rf(x_{n+1})\), then it is necessary to simulate
\(\rf(x_{n+1})\) \textbf{conditional on} \(\rf(x_1), \dots, \rf(x_n)\).

Theorem~\ref{thm: conditional gaussian distribution} about conditional Gaussian
distributions is well known
\citep[e.g.][Prop.~3.13.]{eatonMultivariateStatisticsVector2007}. But in the
proof we will see the relation to the cholesky decomposition.

\begin{theorem}[Conditional Gaussian distribution]
    \label{thm: conditional gaussian distribution}
    Let \(X\sim\normal(\mu,\Sigma)\) be a multivariate Gaussian vector where
    the covariance matrix is a block matrix of the form
    \[
        \mu = \begin{bmatrix}
            \mu_1\\ \mu_2
        \end{bmatrix}
        \quad \text{and} \quad
        \Sigma = \begin{bmatrix}
            \Sigma_{11} & \Sigma_{12}
            \\
            \Sigma_{21} & \Sigma_{22}
        \end{bmatrix},
    \]
    then the conditional distribution of \(X_2\) given \(X_1\) is
    \[
        X_2\mid X_1 \sim \normal(\mu_{2\mid 1}, \Sigma_{2\mid 1}),
    \]
    with conditional mean and variance
    \begin{align*}
        \mu_{2\mid 1} &:= \mu_2 + \Sigma_{21}\Sigma_{11}^{-1}(X_1-\mu_1)
        \\
        \Sigma_{2\mid 1} &:= \Sigma_{22} -  \Sigma_{21}\Sigma_{11}^{-1}\Sigma_{12}.
    \end{align*}
    where \(\Sigma_{11}^{-1}\) may be a generalized inverse
    \citep[cf.][]{eatonMultivariateStatisticsVector2007}.
\end{theorem}
\begin{proof}
    For the general statement we point to the standard literature
    \citep[e.g.][]{eatonMultivariateStatisticsVector2007}.
    For this proof we will assume for simplicity that \(\Sigma\) is invertible.

    Let \(\bar{X} := X-\mu\) be the centered version of \(X\). Then there exists
    a unique lower triangular matrix \(L\) such that \(\Sigma = LL^T\) (i.e.\ the Cholesky
    Decomposition). This results in the following representation\footnote{
        It is straightforward to check that \(Y := L^{-1}(X-\mu)\)
        is a multivariate Gaussian vector of centered, uncorrelated entries with variance \(1\).
        They are therefore \(\iid\) standard normal since uncorrelated multivariate Gaussian
        random variables are independent. Sometimes this representation is even
        taken as the definition of a multivariate normal distribution.
        This is the only real place we use the invertibility of \(\Sigma\) in
        its entirety, the invertibility of \(\Sigma_{11}^{-1}\) is used later on
        because we do not want to get into the business of generalized inverses
        here.
    }
    \[
        X - \mu =:
        \begin{bmatrix}
            \bar{X}_1
            \\
            \bar{X}_2
        \end{bmatrix}
        = \begin{bmatrix}
            L_{11} & 0
            \\
            L_{21} & L_{22}
        \end{bmatrix}
        \begin{bmatrix}
            Y_1 \\ Y_2
        \end{bmatrix}
        = LY
    \]
    with independent standard normal \(Y_i\), i.e. \(Y\sim\normal(0, \identity)\).
    \(L_{11}\) is invertible since \(\Sigma_{11}\) is invertible and therefore the map from
    \(Y_1\) to \(X_1\) is bijective. Conditioning on \(X_1\) is therefore
    equivalent to conditioning on \(Y_1\). But we have
    \begin{equation}
        \label{eq: underlying explicit decomposition}
        X_2 = \mu_2 + \bar{X}_2
        = \underbrace{\mu_2 + L_{21}Y_1}_{\text{conditional expectation}} + \underbrace{L_{22} Y_2}_{\text{conditional distribution}}
    \end{equation}
    So it follows that
    \[
        X_2\mid X_1 \sim  \normal(\mu_{2\mid 1}, \Sigma_{2\mid 1})
    \]
    with
    \begin{align*}
        \mu_{2\mid 1} &:= \E[X_2 \mid X_1]
        &=& \mu_2 + L_{21}Y_1
        \\
        \Sigma_{2\mid 1} &:= \Cov[X_2 \mid X_1] = \E\Bigl[ \bigl(X_2 - \E[X_2 \mid X_1]\bigr)^2 \mid X_1\Bigr]
        &=& L_{22}L_{22}^T.
    \end{align*}
    What is left to do, is to find a representation for the \(L_{ij}\)
    using the block matrices of \(\Sigma\). For this we observe
    \begin{equation}
        \label{eq: sigma to L conversion}
        \setlength\arraycolsep{4pt}
        \begin{bmatrix}
            \Sigma_{11} & \Sigma_{12}\\
            \Sigma_{21} & \Sigma_{22}
        \end{bmatrix}
        = \Sigma
        = LL^T = \begin{bmatrix}
            L_{11}L_{11}^T & L_{11}L_{21}^T
            \\
            L_{21}L_{11}^T & L_{22}L_{22}^T + L_{21}L_{21}^T
        \end{bmatrix}.
    \end{equation}
    Using \(Y_1 = L_{11}^{-1}\bar{X}_1\) and the insertion of an identity matrix this implies
    \[
        L_{21}Y_1 = L_{21}(L_{11}^T L_{11}^{-T})(L_{11}^{-1}\bar{X}_1)
        \overset{\eqref{eq: sigma to L conversion}}= \Sigma_{21}\Sigma_{11}^{-1}(X_1-\mu_1).
    \]
    resulting in the desired conditional expectation \(\mu_{2\mid 1}\). The conditional variance follows alike
    \begin{align*}
        \Sigma_{2\mid 1} = L_{22}L_{22}^T
        \overset{\eqref{eq: sigma to L conversion}}&= \Sigma_{22} - L_{21}L_{21}^T
        \\
        &= \Sigma_{22} - \underbrace{L_{21}(L_{11}^T}_{\overset{\eqref{eq: sigma to L conversion}}=\Sigma_{21}}\underbrace{L_{11}^{-T}) (L_{11}^{-1}}_{\overset{\eqref{eq: sigma to L conversion}}=\Sigma_{11}^{-1}}\underbrace{L_{11})L_{21}^T}_{\overset{\eqref{eq: sigma to L conversion}}=\Sigma_{12}}.
    \end{align*}
\end{proof}

\textcolor{red}{Takeaways from the proof:} For conditioning, the cholesky decomposition
is quite natural and the expressions for the conditional expectation and conditional
covariance are shorter using the Cholesky decomposition \(L\) than using the covariance
matrix \(\Sigma\).

\begin{enumerate}
    \item \(L_{21}\) is sort of the ``natural conditional expectation'' (see \texttt{natural\_ce} in the code)
    \item \(L_{22}\) is essentially the ``residual standard deviation''
    that remains after you subtract the explained variance \(L_{21}L_{21}^\transpose\)
    from the unconditional variance \(\Sigma_{22}\) resulting in the conditional
    variance
    \[
        \Sigma_{2\mid 1} = L_{22}L_{22}^\transpose = \underbrace{\Sigma_{22}}_{\text{unconditial variance}} - \underbrace{L_{21}L_{21}^\transpose}_{\text{explained variance}}.
    \]
    \item \(Y_i\) is the underlying \(\iid\) randomness. In the conditional simulation
    we keep the existing \(Y_i\) that we have and add ``new randomness'' (this is
    how it is called in the code)

    \item Observe for the above to work we do not need \(L_{11}\) and \(L_{22}\)
    to be triangular. We only need the upper right corner of \(L\) to be zero.
    For the calculation of \(L_{22}\) from \(\Sigma_{22}\) and \(L_{21}\)
    we can therefore use a more stable algorithm than Cholesky!
    \textcolor{red}{This is not yet implemented, at the moment I am only using
    cholesky in the code. But this would potentially make the algorithm much
    more stable.}

    \item When does cholesky get unstable? When the explained variance is almost
    the same as the unconditional variance, then it is likely that the ``residual variance''
    is numerically negative and we cannot calculate the ``residual standard deviation''
    \(L_{22}\). This happens when the random variables are highly correlated.
    So in case of a random function typically, when the evaluation locations are
    very close to each other.
\end{enumerate}

\subsection{Incremental simulation IS the cholesky decomposition}

Another key point here is that the cholesky decomposition is implemented
exactly like this:

Assuming we already have the decomposition of \(\Sigma_{11}\)
as \(\Sigma_{11} = L_{11}L_{11}^\transpose\), then we can obtain
the cholesky decomposition \(L\) of
\[
    \Sigma = \begin{bmatrix}
        \Sigma_{11} & \Sigma_{12}
        \\
        \Sigma_{21} & \Sigma_{22}
    \end{bmatrix}
    = L L^\transpose
    = \begin{bmatrix}
        L_{11} & 0
        \\
        L_{21} & L_{22}
    \end{bmatrix}
    \begin{bmatrix}
        L_{11}^\transpose & L_{21}^\transpose
        \\
        0 & L_{22}^\transpose
    \end{bmatrix}
    = \begin{bmatrix}
        L_{11}L_{11}^\transpose & L_{11}L_{21}^\transpose
        \\
        L_{11}^\transpose L_{21} & L_{22}L_{22}^\transpose + L_{21}L_{21}^\transpose
    \end{bmatrix}
\]
by calculating
\[
    L_{21} = (L_{11}^{-1}\Sigma_{21})^{-1}
\]
and finally,
\[
    L_{22} = \sqrt{\Sigma_{22} - L_{21}L_{21}^\transpose}.
\]
The square root above is a normal square root if we only add a single row
and \(\Sigma_{22}\) is therefore a \(1\times 1\) matrix.
This is precisely how the Cholesky-Banachiewicz
algorithm for calculating the cholesky decomposition is defined.
You start with \(1\times 1\) matrix and then add a single row in every
iteration.

So if we incrementially simulate \(\rf(x_1), \dots, \rf(x_n)\)
we pay the exact same price as simulating this vector all at once, because
if we simulate it all at once we have to find the cholesky decomposition
of the covariance matrix of this vector, which we also do when we incrementally
simulate.

\textbf{However:} While it is true that we pay no price in computational complexity,
the actual implementation can suffer:
\begin{enumerate}
    \item We cannot use existing BLAS routines, and manual implementation is slower
    \item The BLAS routines progress column by column instead of row by row -- most likely
    because this results in better memory access patterns
    \item We do not know the final size of the covariance matrix, so we need to be able
    to add rows incrementally. Since each row also gets longer in every iteration and we
    do not want to move all of this around in memory this forces a packed memory layout 
    that is not as suited to existing BLAS routines.
\end{enumerate}

Consequently, the above leaves room for a lot of manual optimization as one cannot
rely as much on existing routines.

\textcolor{red}{Observe that if we only add a single random variable \(\rf(x_{n+1})\)
in every iteration, then we only add a single row to the covariance matrix and
this corresponds exactly to the cholesky decomposition. But if we add more than
one random variable at once, then we can use a different factorization algorithm
than cholesky to calculate \(L_{22}\). In particular, if the user asks for all the points
they are interested in right from the start, then we can avoid cholesky entirely.
}
